---
title: "Baseball Playoff Series Information"
author: "Angelica Martinez & Gabriel Rodriguez"
date: "4/14/2022"
output: html_document
---

# Introduction




```{r}

# Libraries we will be using for this exploration

library(tibble)
library(tidyverse)
library(ggplot2)
library(Lahman)

```


## Postseason Batting Data

Before we can use the 'BattingPost' data table we need clean it up a bit. First we want to remove any missing data. There is no 
data available for the 'CS', 'HBP', 'SH', 'SF', and 'GIDP' columns for years 1884-1892, so we will remove data for these years. 
We will also rename the 'round' column to 'Playoff_Series' to make it more understandable.

Second, we will add more relevant baseball calculations that will help us better analyze batting performance.

Finally we will take a look at the means of those statistic for all teams, for each individual playoff series from 2012-2019. 


```{r}

# The 'BattingPost' data frame contains batting information for each MLB postseason game for years
# 1884-2020.
view(BattingPost)
BattingPost <-BattingPost

# Let's take a look at summarized statistics for each different playoff series

# Remove all observations for which data is missing. Rename 'round' column to 'Playoff Series'
PostSeries_Batting <- BattingPost %>% 
  na.exclude() %>% 
  rename("Playoff_Series" = round)

```

This table already contains a lot of relevant baseball statistics. However it is missing important calculations that are commonly
used to determine batting performance.

Fortunately, the data needed to perform these calculations is already in the table, so all we have to do is add columns that perform
these calculations. More specifically, we will want to look at batting average (AVG), on-base percentage (OBP), slugging (SLG), and
on-base plus slugging (OPS).

```{r}
# Add columns that calculate more batting statistics
PostSeries_Batting <- PostSeries_Batting %>% 
  filter(AB > 0) %>%                                              # We only want batting info for players who had at least one AB
  mutate(X1B = (H - (X2B + X3B + HR)),                            # Add a column for singles hit
         TB = (1*X1B) + (2*X2B) + (3*X3B) + (4*HR),               # Add a column for totatl bases
         AVG = H/AB,                                              # Add a batting average column
         OBP = (H + HBP + BB + IBB) / (AB + BB + IBB + HBP + SF), # Add an on-base percentage column
         SLG = TB / AB,                                           # Add a slugging percentage column
         OPS = OBP + SLG)                                         # Add an on-base plus slugging column

         
         
# Remove base running statistics. We won't need those for batting statistics purposes
PostSeries_Batting <- PostSeries_Batting %>% select(-SB, -CS)

# Rearrange columns
PostSeries_Batting <- PostSeries_Batting %>%
  relocate(TB, AVG, OBP, SLG, OPS, .after = BB)

```


**Overall means for batting average (AVG), on-base percentage (OBP), slugging (SLG),** 
**and on-base plus slugging (OPS), for each individual postseason series during the** 
**expanded wild card era (2012-2019)**
```{r}
# Let's focus on the expanded wild card era years 2012-2019
# Compare the batting average among each playoff series

# This function takes in our 'PostSeries_Batting' df, groups_by 'Playoff_Series', then
# summarizes the mean of the chosen stat for each different playoff series, within the 
# chosen start_year and end_year

Stat_Mean <- function(dataset, Playoff_Series, stat_name, start_year, end_year) {
  Playoff_Series <- enquo(Playoff_Series)
  stat_name <- enquo(stat_name)
  
  dataset %>%
  filter(yearID >= start_year & yearID <= end_year) %>%
  group_by(!!Playoff_Series) %>%
  summarize(series_mean = mean(!!stat_name)) %>%
  arrange(-series_mean)
}
  
# Use the 'Stat_Mean' function to calculate the mean AVG, OBP, SLG, and OPS
# for each playoff series for years 2012-2019
Series_AVG <- Stat_Mean(PostSeries_Batting, Playoff_Series, AVG, 2012, 2019)  
Series_OBP <- Stat_Mean(PostSeries_Batting, Playoff_Series, OBP, 2012, 2019)  
Series_SLG <- Stat_Mean(PostSeries_Batting, Playoff_Series, SLG, 2012, 2019)
Series_OPS <- Stat_Mean(PostSeries_Batting, Playoff_Series, OPS, 2012, 2019)
 
```

```{r}
# Plot the results
Series_AVG %>% ggplot(aes(x = reorder(Playoff_Series, -series_mean), y = series_mean)) + geom_col()
Series_OBP %>% ggplot(aes(x = reorder(Playoff_Series, -series_mean), y = series_mean)) + geom_col()
Series_SLG %>% ggplot(aes(x = reorder(Playoff_Series, -series_mean), y = series_mean)) + geom_col()
Series_OPS %>% ggplot(aes(x = reorder(Playoff_Series, -series_mean), y = series_mean)) + geom_col()
```


## Postseason Pitching Data

```{r}

# The 'PitchingPost' data frame contains pitching information for each MLB postseason game for years
# 1884-2020.
view(PitchingPost)
PitchingPost <-PitchingPost

# Let's take a look at summarized statistics for each different playoff series

# Remove all observations for which data is missing. Rename 'round' column to 'Playoff Series'
PostSeries_Pitching <- PitchingPost %>% 
  na.exclude() %>% 
  rename("Playoff_Series" = round)


```


```{r}
# Add columns that calculate more pitching statistics
PostSeries_Pitching <- PostSeries_Pitching %>% 
  mutate(IP = IPouts / 3,                    # Calculate the number of innings pitched
         WHIP = (H + BB + IBB) / IP)         # Calculate the walks/hits per inning pitched

PostSeries_Pitching_2012_2019 <- PostSeries_Pitching %>% 
  filter(yearID >= 2012 & yearID <= 2019)

         

```


```{r}

# Use the stat_mean function to visualize the mean for ERA, IP, and WHIP form 2012-2019

Series_ERA <- Stat_Mean(PostSeries_Pitching, Playoff_Series, ERA, 2012, 2019)
Series_IP <- Stat_Mean(PostSeries_Pitching, Playoff_Series, IP, 2012, 2019)
Series_WHIP <- Stat_Mean(PostSeries_Pitching, Playoff_Series, WHIP, 2012, 2019)


```

